Peer influences in air travel behavior: A social autoregressive analysis

### 2015-02-18 article.Rmd
Limited line length in paragraphs.

Introduction
================================================
The goal of a modern activity-based travel model is to accurately represent travel behavior choices on a microscopic level [@Bhat2003]. As researchers have pursued this goal, they have sought to incorporate individual behavioral patterns of various types. For example, researchers have built models that allow different classes of people to place different values on goods [@Walker2002], that capture idiosyncratic preferences for a particular lifestyle [@Cao2007a], or that allow multiple decisions to be made either sequentially or simultaneously [@Bhat2006; @Pinjari2011].

@@

The distinction between these elements can be subtle. Imagine two individualswho are socially connected to each other. If these individuals meet for dinner, this is a direct influence, because each person’s choice of activity and destination is dependent on the other person being there. On the other hand if one recommends a restaurant to the other, this is an indirect influence because the restaurant entered one’s choice set through the other’s experience and not his own. Finally, if the two know each other through another shared destination or interest (e.g. workplace, neighborhood, church), the restaurant in question may have entered their choice set by being near one of these destinations.

@@

However, both of these situations are changing. Usage of online social networking sites and other social media continues to grow, and contains social connection and preference data for  millions of individuals. On the econometric front, the spatial econometrics literature in the last twenty-plus years [@Anselin2003; @Anselin2011] enables the view of spatial interaction as a special case of social interaction, or as a proxy that may be more easily observed in data.

In this paper, we develop and apply a technique that uses spatial data to analyze airline travel (trip frequency)  of individuals in a common social network. To do this, we construct a series of social weights matrices representing potential relationship strength between members of this network, and incorporate these matrices in a social Durbin model capturing direct, indirect, and endogenous effects. 

@@

>>>>

Introduction
================================================
The goal of a modern activity-based travel model is to accurately represent
travel behavior choices on a microscopic level [@Bhat2003]. As researchers
have pursued this goal, they have sought to incorporate individual behavioral
patterns of various types. For example, researchers have built models that allow
different classes of people to place different values on goods [@Walker2002],
that capture idiosyncratic preferences for a particular lifestyle [@Cao2007a],
or that allow multiple decisions to be made either sequentially or
simultaneously [@Bhat2006; @Pinjari2011].

@@

The distinction between these elements can be subtle. Imagine two individuals who
are socially connected to each other. If these individuals meet for dinner, this
is a direct influence, because each person's choice of activity and destination
is dependent on the other person being there. On the other hand if one
recommends a restaurant to the other, this is an indirect influence because the
restaurant entered one's choice set through the other's experience and not his
own. Finally, if the two know each other through another shared destination or
interest (e.g. workplace, neighborhood, church), the restaurant in question may
have entered their choice set by being near one of these destinations.

@@

However, both of these situations are changing. Usage of online social
networking sites and other social media continues to grow, and contains social
connection and preference data for  millions of individuals. On the econometric
front, the spatial econometrics literature in the last twenty-plus years
[@Anselin2003; @Anselin2011] enables the view of spatial interaction as
a special case of social interaction, or as a proxy that may be more easily
observed in data.

In this paper, we develop and apply a technique that uses spatial data to
analyze airline travel (trip frequency)  of individuals in a common social
network. To do this, we construct a series of social weights matrices
representing potential relationship strength between members of this network,
and incorporate these matrices in a social Durbin model capturing direct,
indirect, and endogenous effects.

@@

Social Networks and Social Econometrics
=======================================================

Social Networks, Online Observation, and Facebook
-------------------------------------------------------
The concept of a social network provides a construct in which the structural and
functional dimensions of individuals’ social relationships (O'Reilly 1988)
can be systematically examined for social influence, or other characteristics of
interest.  Social network structure, which refers to the individuals and
linkages comprising one's social network (Due, Holstein et al. 1999) can be
analyzed either through egocentric network analysis (focused on a single
individual and his relationships with others) or whole network analysis (focused
comprehensively on the relationships among all members in a network).  Social
network function clarifies the nature of, and interpersonal interaction among,
the linkages within the social network structure; for example, "social support,
relational strain, or social anchorage," among others. (Due, Holstein et al.
1999) In the context of this study, one important dimension of social network
function could be the level of influence associated with an individual
connection between two individuals, or among many connections within a group of
individuals.

Data collection, however, complicates this type of study.   The predominant
method employed by previous studies (for example, (27–29)) to construct
models of individuals’ social networks, called “name generators,” uses sets of
interview questions designed to solicit information from an individual regarding
whom he considers as part of his social network (structure) and the role or
importance of that individual (function).  The name generator approach can
become problematic, however, when respondents have difficulty in recalling
information regarding their social networks (15), thus providing information
to the researchers that is unknowingly incomplete.  In addition, the
name-generator approach can be  time consuming, with data collection interviews
frequently taking several hours for a single interview (30) to construct a
model of one individual's egocentric social network.

Online social networking websites, such as Facebook and Twitter, provide an
alternative source for social network data, the collection of which allows for
the delineation of naturally developing social networks with unambiguous
boundaries (thus overcoming respondent recollection issues) (14, 15).
Additionally, as such data is stored in a centralized computer database and in a
format that is consistent across all users, social network data can be collected
in seconds and with greater structural consistency.  Thus, data collection is
far less time consuming and data quality and consistency is potentially improved
as compared to the name generator approach (14, 15).

In this study, online social network data collection is focused exclusively on
users of the social networking site, Facebook. With 864 million daily active
users, and 1.12 billion monthly active users globally as of September 2014,
Facebook is among the world's largest social networking sites (Facebook
2015) (by comparison, Twitter has 284 million monthly active users across the
globe as of October 2014). (Twitter 2015).  Facebook enables users to create
online profiles (e.g., general demographic, educational, special interest, and
geographical information), which is shared with other users.  As users connect
with other individuals, becoming 'friends,' an online model of social
relationships and connections is created. (Wall, Macfarlane et al. 2014). A
study by the Pew Research Center found that the primary motivation for the
majority of adult  users of social networking sites in the United States is
"staying in touch with family members," and "staying in touch with current
friends and reconnecting with old friends." (Smith 2011) It therefore may be
reasonable to "suspect that an individual’s online social network structure
represents, to some extent, an online model of that individual’s real-life
social network structure." (Wall, Macfarlane et al. 2014)

There are also several characteristics of the Facebook dataset, as well as the
types of data contained in that dataset, that are particularly relevant to the
consideration of socio-spatial influences. Lewis et al. (2008) discusses
five defining features of the Facebook social network model:

	1. Naturally Occurring:  Facebook data is naturally occurring when downloaded
     directly from Facebook (i.e. not subject to a respondent recollection ability) 
	2. Whole Networks: Facebook.com contains complete network data of a
     sociocentric nature, as opposed to an egocentric nature, which reflects the
     interconnectedness of network members
	3. Longitudinal Data: Facebook data may reflect how relationships grow and
     change over time; 
	4. Cultural Insight: Facebook users can indicate tastes and values, which may
     be reflected in the data 
	5. Relationship Strength: Facebook can provide data on relationship strength.

Several of these items, as well as the types of data contained in Facebook
datasets are particularly relevant to the consideration of socio-spatial
influences.

The whole network aspect of Facebook data enables an analysis of interconnected
or shared relationships among a population.  Further, relationship strength
provides some insight into the function of relationships, as may be related to
the level of indirect influence across a relationship.  Lastly, and very
importantly, Facebook maintains locational information that can easily be
translated to geospatial

Spatial Econometrics
---------------------------------------
Spatial econometrics is built on the understanding that observations near to
each other spatially are likely to exhibit correlation and interdependence, as
they share common unobserved variables and influence each other. A class of
models developed to solve these problems was pioneered by Anselin (1988);
LeSage & Pace (2009) provide a full and modern treatment.

For a linear regression model , a necessary assumption is that the observations
are independent, which is unlikely in spatial data. Tobler’s law implies that if
two observations are “close” to each other, then they are likely to share common
characteristics (Tobler 1970). Further, observations close to each other may
actually exert an influence on each other in the way air pollution lowers the
price of nearby properties (Kim & Goldsmith 2008). If correlation and
interdependence are not controlled in an econometric model, then the model may
return biased parameters and/or invalid hypothesis tests.

This correlation and interdependence can be accommodated with a spatial Durbin
model (SDM),

where  is a spatial weights matrix in which each element  maps the spatial
relationship between two observations  and ; if  and  are “neighbors,” then .
The parameter  measures the dependence between  and , and the lagged independent
variable coefficients in  contain the impact that attributes of  have on .

Such spatial models have been applied to numerous different contexts, including
transportation modeling and forecasting. In most if not all of these cases,
however, spatial data is used as a proxy for an inferred social relationship.
For example, Adjemian et al. (2010) observed that vehicle body types were
spatially autocorrelated, with the authors inferring that neighbors may
influence each other in choosing vehicle types. Similarly, Bhat et al.
(2010) used spatial proximity as a means to study the social destinations of
teenagers. In these and other cases, it may be more appropriate to use the
social network directly, if the social network can be translated to a  matrix.
This is effectively the approach of Bramoullé et al. (2009) and
Calvó-Armengol et al, (2009), who showed that a student's school performance
and physical activity level (respectively) are socially interdependent: students
with studious or athletic friends tend to be so as well.

There are two primary difficulties in creating , one econometric and one
practical. From an econometric standpoint, the researcher typically must specify
exogenously. Though some common practices have emerged in the spatial literature
(Dubin 1998), not all of these can be readily adopted to social networks.
This gap in the literature is one that we seek to fill with this study, by
examining several possible  matrices that we construct from social network data.
The practical difficulty is actually collecting that data, though this is
becoming less difficult as interaction within social networks happens through
online sites such as Facebook. We discuss this practical evolution in Section
2.2.

Methodology
======================================
We introduce social dependence into a Poisson count model that predicts the
number of airline trips an individual makes as a function of his or her
socioeconomic variables. The Poisson count model estimates the effect of
dependent variables  on the conditional mean number of trips, . We consider
social interaction in this model by suggesting that  (where  is the row of  that
contains the weights for the neighbors of ). Our goal in this analysis is to
recover a weighting scheme  that maximizes the model likelihood, and use this
model to examine the inferential error that could result from ignoring social
interdependence and correlation in such model.

Data Collection
----------------------------------------
To examine the potential for influences to travel behavior between individuals’
within an interconnected social network, a web-based survey was developed, and
respondents were recruited through the Georgia Tech fraternity and sorority
system (a social network that may presumably have a proportionally higher level
of interconnectedness), in partnership with the Georgia Tech Dean of Students,
Office of Greek Affairs.  The web-based survey instrument was distributed via
email to the primary points of contact for each fraternity and sorority on file
at the Office of Greek Affairs.  Fraternities and sororities were asked to also
encouraged to solicit responses from their alumni, thus, potentially increasing
the sample population of approximatley 3,300 students in Greek housing.
Fraternities and sororities were incentivized with the promise of a $500
charitable contribution made on behalf of the fraternity and the sorority each
with the highest participation.

The online survey instrument was adapted from Wall et al (2014), and
contains three principal components.  For a detailed description of the
variables collected, refer to Wall et al. (2014).  The first component
collected respondents’ air travel information from the past year – this includes
a list of trips taken with origin and destination airports for round-trip,
multi-destination, and one-way travel.  The second component collected
socio-demographic information, including age, gender, income, education level,
citizenship, occupation, and race.  The third component asked respondents to log
in to their Facebook accounts, and the survey automatically collected data
contained in the respondent's profile and friends list via the Facebook
application programming interface (API).  This data included the hometown city
and current city locations of the respondents, and also that of the respondents’
friends.  The program also collected an encrypted (one way hash) version of the
Facebook ID for all respondents and their friends.  This ensured subject
anonymity while simultaneously allowing observations regarding 1) which
respondents were friends with each other, and 2) how many friends respondents
have in common, 3) identifying and limiting the potential for duplicate
respondents.

Because the survey program collects Facebook profile and friends list data from
the perspective of the respondent's individual profile, it is limited in its
data collection to "visible" user profile and friend data. That is, individuals'
information that was restricted as a result of more stringent individual privacy
settings would not be visible to the survey program. A consequence of this is
potential for imbalance in privacy settings (and thus, data visibility) between
two connections, is that friend connections may be  observed as one-way in the
data even though one-way friend connections are not possible on Facebook.

Facebook locational data (i.e., hometown and current city location) is not
geocoded, and therefore required some post-processing prior to any analyses.
Facebook location data were post-processed using R statistical programming
language and ArcGIS 10.  The study analyzed respondents’ friends who do not
consider Atlanta, Georgia as their hometown.  Friends’ hometown cities were
geocoded as point locations. To associate cities as within (or outside of) the
Atlanta metropolitan region, circular buffers were created (using
Hartsfield-Jackson Atlanta International airport as the centroid) at 25 mile
radius increments from 25 miles to 150 miles. The 6 resulting hometown buffer
regions were then associated with the geocoded city locations, and thus the
number of friends with hometown city locations within each buffer region were
determined. Only data from respondents that completed the entire survey (i.e.
the air travel diary, socio-demographic section, and the Facebook data
collection) were considered for the analysis.

Candidate Weighting Schema
-----------------------------------------
The social network data we retrieved from Facebook allow us to construct four
social weights matrices, which we describe in this section. To aid this
presentation, we must first establish some basic notation following elementary
set theory. The set of friends of an individual  is defined as . The cardinality
of this set is the number of friends, so therefore the quantity  represents the
number of friends that individuals  and  have in common.

**Direct Friendship**
This schema defines a simple binary matrix where a cell is nonzero if  and  are
friends. Specifically,

**Mutual Friendship**
This schema defines a matrix where the cell value is the number of friends  and
have in common (as in Gilbert & Karahalios 2009; He et al. 2012),

**Network Overlap**
This schema defines a matrix where the cell value is the relative overlap of
each  pair’s social network (as in Adamic & Adar 2003),

**Direct Network Overlap**
This schema is similar to the Network Overlap, except that we give a bonus to
pairs that are themselves friends,

By Facebook’s internal logic, each of these matrices must be symmetric: it is
not possible on Facebook to have non-reciprocal friends. However, the individual
user privacy settings make it possible that we would observe apparent one-way
relationships through the API; indeed, this affected approximately 10% of cases.
We therefore coerce this and other schema to be symmetric as appropriate.

As a final note on the schema development, behavioral inference on and
estimation of the models is simplified by using a row-standardized matrix, where
the rows sum to unity(LeSage & Pace 2009, p.85). We follow this convention
in all cases.

Results and Discussion
================================

Conclusions
================================

# References {-}
